---
title: "Populism in Pan-American Politics"
author: "Jackson R. Mandel"
date: "2025-04-09"
output:
  beamer_presentation:
    theme: Madrid
    colortheme: orchid
    fonttheme: structuresmallcapsserif
  slidy_presentation: default
header-includes: \usepackage{palatino}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#Setup
library(tidyverse)
library(dplyr)
library(ggplot2)
library(countrycode)

#Download Data
fulldata <- readRDS("V-Dem-CPD-Party-V2.rds")

#Filter Data
filtereddata <- fulldata %>%
  select(v2paenname, country_name, year, v2xpa_popul, v2paseatshare)

#Clean Data & Remove Parties with less than 10% of seats in legislature
cleandata <- filtereddata %>%
  filter(!is.na(v2paseatshare) & !is.na(v2xpa_popul) & v2paseatshare >= 10)

#Regression
PopulismRegression <- lm(v2xpa_popul ~ year, data = cleandata)
print(PopulismRegression)
ggplot(cleandata, aes(x = year, y = v2xpa_popul)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Scatterplot of Populism Level vs. Year", 
       x = "Year", 
       y = "Populism") +
  theme_minimal()

#Caluclating Country Averages
average_populism <- cleandata %>%
  group_by(country_name, year) %>%
  summarize(average_populism = mean(v2xpa_popul, na.rm = TRUE))%>%
  group_by(country_name) %>%
  filter(year == max(year)) %>%
  ungroup()

#Ranking countries by level of populist rhetoric in their political parties
ranked_populism <- average_populism %>%
  mutate(populism_rank = rank(-average_populism, ties.method = "first"))

#Boxplot of Country Averages
ggplot(ranked_populism, aes(y = average_populism)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  labs(title = "Distribution of Average Populism Rankings", 
       y = "Average Populism Score") +
  theme_minimal()

#Creating a variable for continent
bycontinent <- ranked_populism %>%
  mutate(continent = countrycode(country_name, origin = "country.name", 
                                 destination = "continent"),
         continent = ifelse(continent == "Americas", "America", continent))
```

## Introduction

- Central Question: Does the average viable political party in the Americas engage in more populist rhetoric than the average viable political party in other countries?
- Motivation:
     - Understanding political dynamics unique to the Americas
     - Uncovering structural similarities between Anglo-American and Latin American political systems
- Conclusion: Although the average populism score of pan-American parties was higher in some regions of the world, the difference was not statistically significant.

## Data

- Drawn from the V-Dem (Varieties of Democracy)V-Party database
- v2xpa_popul: a measure of populist rhetoric by a political party
  - Aggregate of a score of anti-elite rhetoric and rhetoric identifying "the people" as a homogenous sovereign interest group
- v2paseatshare: the share of seats in parliament at the last election
     - Used to filter out parties with less than 10% of seats
- country_name: used to create a new variable identifying what continent a country is on.

## Comparing Populist Rhetoric

```{r boxplot}
#Americas or Not Boxplot
ggplot(bycontinent, aes(x = factor(continent, levels = c("America", "Africa", "Asia", "Europe", "Oceania", "Other")), 
                         y = average_populism)) + 
  geom_boxplot(fill = "royalblue", color = "black") + 
  labs(title = "Average Populism: Americas & Rest of World", x = "", 
       y = "Average Populism Score") + 
  theme(
    plot.title = element_text(size = 20, face = "bold", color = "black", family = "serif"),
    axis.title = element_text(size = 18, face = "bold", color = "black", family = "serif"),
    axis.text = element_text(size = 14, face = "bold", color = "black", family = "serif"),
    axis.line = element_line(color = "black"),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

## Success of Populist Parties in the Americas

```{r regression, warning=FALSE, message=FALSE, fig.width=6, fig.height=4}
americasdata <- cleandata %>%
  mutate(continent = countrycode(country_name, origin = "country.name", destination = "continent")) %>%
  filter(continent == "Americas")
ggplot(americasdata, aes(x = v2paseatshare, y = v2xpa_popul)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(
    title = "Regression of Populism on Seat Share in the Americas",
    x = "Seat Share (%)",
    y = "Populism Index"
  ) +
  theme(
    plot.title = element_text(size = 18, face = "bold", color = "black", family = "serif"),
    axis.title = element_text(size = 16, face = "bold", color = "black", family = "serif"),
    axis.text = element_text(size = 12, face = "bold", color = "black", family = "serif"),
    axis.line = element_line(color = "black"),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

