---
title: "Populism in Pan-American Politics"
author: "Jackson R. Mandel"
date: "2025-04-09"
output:
  beamer_presentation:
    theme: Madrid
    colortheme: orchid
    fonttheme: structuresmallcapsserif
  slidy_presentation: default
header-includes: \usepackage{palatino}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#Setup
library(tidyverse)
library(dplyr)
library(ggplot2)
library(kableExtra)

#Download Data
fulldata <- readRDS("V-Dem-CPD-Party-V2.rds")

#Filter Data
filtereddata <- fulldata %>%
  select(v2paenname, country_name, year, v2xpa_popul, v2paseatshare)

#Clean Data & Remove Parties with less than 10% of seats in legislature
cleandata <- filtereddata %>%
  filter(!is.na(v2paseatshare) & !is.na(v2xpa_popul) & v2paseatshare >= 10)

#Regression
PopulismRegression <- lm(v2xpa_popul ~ year, data = cleandata)
print(PopulismRegression)
ggplot(cleandata, aes(x = year, y = v2xpa_popul)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Scatterplot of Populism Level vs. Year", 
       x = "Year", 
       y = "Populism") +
  theme_minimal()

#Caluclating Country Averages
average_populism <- cleandata %>%
  group_by(country_name, year) %>%
  summarize(average_populism = mean(v2xpa_popul, na.rm = TRUE))%>%
  group_by(country_name) %>%
  filter(year == max(year)) %>%
  ungroup()

#Ranking countries by level of populist rhetoric in their political parties
ranked_populism <- average_populism %>%
  mutate(populism_rank = rank(-average_populism, ties.method = "first"))

#Boxplot of Country Averages
ggplot(ranked_populism, aes(y = average_populism)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  labs(title = "Distribution of Average Populism Rankings", 
       y = "Average Populism Score") +
  theme_minimal()

#Creating a binary variable for whether countries are in the Americas

americas_countries <- c("Dominican Republic", "Haiti", "Guyana", "Uruguay", 
                        "Jamaica", "El Salvador", "Panama", "Honduras", 
                        "Argentina", "Barbados", "Canada", "Colombia", 
                        "Trinidad and Tobago", "Ecuador", "Guatemala", 
                        "Paraguay", "Mexico", "Peru", "Nicaragua", "Brazil", 
                        "Costa Rica", "Venezuela", "Cuba", 
                        "United States of America")
americasdata <- ranked_populism %>%
  mutate(is_americas = ifelse(country_name %in% americas_countries, 
                              "Americas", "Other"))
```

## Introduction

- Central Question: Does the average viable political party in the Americas engage in more populist rhetoric than the average viable political party in other countries?
- Motivation:
     - Understanding political dynamics unique to the Americas
     - Uncovering structural similarities between Anglo-American and Latin American political systems
- Conclusion: Although the average populism score of pan-American parties was higher than the rest of the world, the difference was not statistically significant.

## Data

- Drawn from the V-Dem (Varieties of Democracy)V-Party database
- v2xpa_popul: a measure of populist rhetoric by a political party
  - Aggregate of a score of anti-elite rhetoric and rhetoric identifying "the people" as a homogenous sovereign interest group
- v2paseatshare: the share of seats in parliament at the last election
     - Used to filter out parties with less than 10% of seats
- country_name: used to create a new variable identifying if a country is in the Americas

## Comparing Populist Rhetoric

```{r boxplot}
#Americas or Not Boxplot
ggplot(americasdata, aes(x = factor(is_americas, labels = 
                                      c("Rest of World", "Americas")), 
                         y = average_populism)) +
  geom_boxplot(fill = "royalblue", color = "black") +
  labs(title = "Average Populism: Americas & Rest of World", x = "", 
       y = "Average Populism Score") +
  theme(
    plot.title = element_text(size = 20, face = "bold", color = "black", family = "serif"),
    axis.title = element_text(size = 18, face = "bold", color = "black", family = "serif"),
    axis.text = element_text(size = 14, face = "bold", color = "black", family = "serif"),
    axis.line = element_line(color = "black"),
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )
```

## Difference of Means Test
- Null Hypothesis: The average viable political party in the Americas does not use more populist rhetoric than the average political party in the rest of the world.
  - Result: failure to reject the null hypothesis
```{r t test}

#Perform a t-test
t_test_result <- t.test(average_populism ~ is_americas, 
                        data = americasdata %>% 
                          filter(!is.na(average_populism)))

cat("\\scriptsize", capture.output(t_test_result), sep = "\n")
```
